Gompertz growth as infinitely correlated non-local growth

### 2024-03-08 arxiv2/new_draft.tex
finalize second draft

It is worth noting that the notion that Gompertz growth is maximally coherent and non-local has also been derived from quantum mechanical principles \cite{molski2003coherent}. Molski and Konarski (2003) showed that the Gompertz equation is the solution of a form of the time-independent Schrodinger equation. From this result, they derive the quantum coherent states which are non-local in space while moving along a classical time trajectory. This adds to our exhibits that the Gompertz model can be interpreted as non-local in origin. Coherent quantum states was originally formulated in the context of electromagnetic fields by \citet{glauber1963coherent} (for which he received the Nobel Prize). Perhaps some biological macroscopic growth phenomena are also due to such fields. For example, the role of internal electrical fields in regenerative growth is well established and is based on negative electrical potentials that act on sensitive cells \cite{becker1984electromagnetic}.

@@

On the topic of disease modelling, the Gompertz model has been shown by Wang et al. \cite{wang2012richards} to be excluded from a subset of models in the communicable disease paradigm. They used the common Susceptible-Infected-Recovered (SIR) model by \citet{kermack1927contribution} augmented by the Richards parameter and showed that the allowed parameter range with the communicable disease assumption does not cover the parameter value taken in the Gompertz scenario. If one incorporates a social proximity network on which the disease spreads, \citet{zonta2022virus} observed through computer simulations that a near Gompertz growth can occur but only for scale-free networks, a type of network that has been shown to be rare in human social networks \cite{clauset2009power,broido2019scale,holme2019rare}. 
We will show that one can also obtain a Gompertz curve from a ubiquitous stressor scenario based on microscopic interactions.
% Given these observations, any mortality pattern that shows a Gompertz characteristic seems not be a result of a communicable disease, but rather from a ubiquitous stressor. 
While many studies show the SIR model (and its extensions) as the fitted model in epidemics \cite{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}, some recent studies have successfully fitted a Gompertz model \cite{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}.

\subsection{Main Results and Roadmap}
In this paper, our primary aim is to investigate the source of growth in self-organized processes. Second, we will elaborate on the idea that the logistic model results from local interactions while the Gompertz model results from a field or equivalently from an infinite set of higher order interactions between the microscopic entities involved in the growth. We will devise a slightly different parametrization than the Richards model, the generalized Gompertz-Logistic equation which is commonly used to compare the two models (e.g. by \citet{petroni2020logistic}, \citet{tjorve2017use}, and \citet{wang2012richards}). In particular, instead of considering a decimal number as the unifying exponent we will appeal to integer exponents with a number of terms that tend towards infinity as we approach the Gompertz model. As we shall see, this is rooted in the microscopic interpretation of the growth system, where each subgroup of $n$ elementary entities interacts with all other equally sized subgroups each composed of mutually exclusive elementary entities. If these subgroups are present at all orders of cardinality, the Gompertz model emerges, while only pairwise interactions yield the logistic model. As seen in the aforementioned references, this is in line both with experimental and theoretical results. In presenting this argument, we present a novel bridge between the logistic and the Gompertz model which offers both a mathematical and biological foundation to the source of growth. Lastly, we will present a unifying framework in the growth context for the common functions Gaussian, Exponential, Logistic, Gompertz and Poisson.

Our paper is structured as follows: we will first present a new macroscopic foundation of the Gompertz-Logistic spectrum which will offer an alternative to the traditional Richards model. Then, we present a microscopic foundation which acts as the building blocks of the macroscopic phenomena. This is where we show that the Gompertz model results from an infinite set of higher order interactions between the elementary entities involved in the growth, which could alternatively be seen as a ubiquitous field or in the biological context: an environmental stressor. We proceed with a discussion that summarizes our findings in the context of the other well-known growth modalities involving the Gaussian and the Laplacian model and then conclude.

@@

\subsection{Microscopic Logistic Growth}
Consider $N$ entities each represented by a set of independent and identically distributed random abundance variables, $x_i(t), \quad i=1, ..., N$, which are unitless and take values between 0 and 1, where 0 represents no realized growth while 1 represents the maximum potential growth having been realized and $\mu$ representing the mean of the distribution. With these variables it will also be convenient to also define an \textit{abundance potential} variable, $s_i = 1 - x_i$, where we omit the temporal variable for ease of exposition.

Now, consider a network between these $N$ entities along a graph with a set of edges and nodes which delineates the interaction of the growth variables, where we define interaction in this context as the transfer of one entity's potential growth to another entity which converts potential to realized growth. Then consider the growth equation with only pairwise interactions, or \textit{first order} interactions, governed by an adjacency matrix ${a_{ij}}$ as a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise, viz.
\begin{equation}
\label{eq:modSIR}
\frac{d x_i}{dt} = x_i\sum_{j \in \mathcal{N}(i)}{a_{ij}}s_{j} \quad \forall i,
\end{equation}
where $\mathcal{N}(i)$ is the neighborhood of nodes connected to node $i$.
First note that for a single pair this would amount to traditional logistic growth. We choose to work in a unitless domain without parameters for the interaction strength and carrying capacity for simplicity of exposition. These can be added without changing the conclusions of these arguments.

Now as we let N grow, we impose a simple rule: \textit{Each entity can only receive growth potential from a single entity and create growth potential in a single possibly different entity}. Using this rule and rearranging and summing across all entities,
\begin{equation}
\label{eq:modSIR2}
\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ] = \sum_j^N s_{j}.
\end{equation}

Dividing by the total number of entities and with our assumption of the variables being independent and identically distributed, we obtain the simplest form of logistic growth of the geometric mean of the ensemble driven by the sample mean of the inverse abundance, or the \textit{potential abundance}, 
\begin{equation}
\label{eq:modSIR3}
\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} = \frac{1}{N}\sum_j^N s_j 
\end{equation}

As N grows, both of the geometric and the arithmetic means converge in expectation to their respective distributional parameters, and thus we can identify these macroscopic quantites starting from pairwise microscopic principles. Specifically, letting $\tilde{x}(N)$ represent the sample geometric mean of $x_i$ and $\bar{x}(N)$ represent the sample arithmetic mean, each converging to their distributional counterparts $\gamma$ and $\mu$, we have 
\begin{eqnarray*}
  \lim_{N\rightarrow \infty} \frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} &=& \lim_{N\rightarrow \infty} \frac{1}{N}\sum_i^N s_i \\\
  \lim_{N\rightarrow \infty} \frac{d}{dt} \ln \tilde{x}(N) &=& \lim_{N\rightarrow \infty}(1 - \bar{x}(N)) \\
  \frac{d}{dt} \ln \gamma &=&(1 - \mu),
  % \frac{d}{dt} \ln m &=&- \ln \bar{s},
\end{eqnarray*}
where the last equality follows from a convergence in expectation.

We make special note that logistic growth emerges from pairwise interactions only. This does not mean that the logistic growth cannot emerge from other microscopic settings but it nevertheless provides one possible scenario under which the logistic curve can emerge. Now, as we shall allow higher order interactions we will see the Gompertz curve emerge.

@@

Thus, let $S_k$ be the set of all mutually exclusive $k$-sized groups where $k=1,..., N/2$, where the highest order are capped at $N/2$ since that would be the largest number of entities in a single group that could interact with an identically sized group. This means that we consider the geometric mean of each $k$-group as follows,
\begin{equation}
\label{eq:modSIR4}
\frac{d}{dt} \ln x_{i_1} x_{i_2} \dots x_{i_k} = s_{j_1}s_{j_2}\dots s_{j_k}, 
\end{equation}
where the indices $i$ and $j$ represent the higher order interaction indices between the mutually exclusive groups of entities for each $k$ from 1 and up to $N/2$. Thus, the aggregated growth is represented by the sum of each interaction order,
\begin{equation}
\label{eq:MicroGomp0}
\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ] = \sum_{j\in S_1}^N s_{j} + \sum_{j_1, j_2 \in S_2} s_{j_1}s_{j_2} + ... + \sum_{j_1, j_2, ..., j_{N/2} \in S_{N/2}} s_{j_1}s_{j_2}\dots s_{j_{N/2}},
\end{equation}
where $S_k$ is the set of all mutually exclusive $k-$sized groups. In each of the sums the total number of terms is equal to the total number of groups which at each order is $N/k$ where $k$ is the interaction order. Thus, if we divide the right hand side by $N$, we can identify a convergence property as both $k$ and $N$ grows:
\begin{enumerate}
\item As the number of entities grow each term will converge in expectation to higher powers of the mean of $s_i$, multiplied by the prefactor $1$, $1/2$, $1/3$, etc., 
\begin{equation}
\label{eq:MicroGomp11}
\lim_{N\rightarrow \infty} \frac{1}{N}\sum_{j_1, j_2, ..., j_k \in S_k} s_{j_1}s_{j_2}\dots s_{j_k} = \frac{1}{N}\frac{N}{k} \mu^{k} = \frac{1}{k} \mu^{k}.
\end{equation}
\item As the higher order interaction terms grow
\begin{equation}
\label{eq:MicroGomp12}
\lim_{k\rightarrow \infty} \frac{1}{N}\sum_{j_1, j_2, ..., j_k \in S_k} s_{j_1}s_{j_2}\dots s_{j_k} = 0
\end{equation}
\end{enumerate}
Thus, as each higher order term is composed of a decreasing number of summations and an increasing number of multiplications, the product of $s_i$ goes to zero since each random variable is $iid$ and bounded between 0 and 1. 
Now, using the Taylor series identity $\mu+\mu^2/2+... = -\ln(1-\mu)$, we obtain Gompertz growth in terms of the arithmetic and the geometric means of the microscopic entities
\begin{eqnarray}
\label{eq:MicroGomp13}
\lim_{N\rightarrow \infty}\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} &=&  \lim_{N\rightarrow \infty}\left[\sum_j^N s_{j} + \sum_{j_1, j_2 \in S_2} s_{j_1}s_{j_2} + ...\right]\\
\frac{d}{dt} \ln \gamma &=&  \mu + \frac{1}{2} \mu^2 + ...\\
\frac{d}{dt} \ln \gamma &=& - \ln (1-\mu).
\end{eqnarray}

Note that for the right hand side to exhibit asymptotically infinite growth rate at the onset of growth, it is not only sufficient but also necessary to have higher order terms. This is seen by virtue of the infinite polynomial representation of the logarithm and implies that Gompertz growth cannot result from pairwise interactions under the assumptions presented in this model, but rather requires all higher order interactions to be present.

Since all higher order interaction would necessarily imply non-local interaction, the natural interpretation would be that the growth is activated by a ubiquitous field accessible by all entities simultaneously. 
In this way one could think of the activating field as an \textit{effectively} external field, even though it is possible that the field emerges as a collective emergent phenomenon purely from within the system or organism.

Another way to see the source of growth as effectively external is to recognize that the source of growth in the higher order terms is coming from the abundance potential and not the realized potential, while in the logistic growth model this is not necessarily the case. 
To see this, first swap the role of $x_i$ and $s_i$ on the right hand side of Equation \ref{eq:modSIR} to see that the same result can be obtained under the exclusivity condition. 
Then note that when augmenting the growth equation with higher order terms on the right-hand side of the same equation, the same swapping will not lead to the Gompertz model.

This causality reversal is indeed what is done when modelling communicable diseases in a microscopic network context, where $x_i$ represents the infected nodes and $s_i$ represents the susceptible nodes (see e.g. \citet{estrada2022networked}). But because higher order growth does not allow this causality reversal, where the infected nodes are the source of spread, the Gompertz model cannot be used since it would imply that the susceptible nodes are the source of the disease spreading, which would turn communicable disease theory on its head. In this context, the Gompertz model would rather suggest a system disturbed by a ubiquitous, simultaneous, and non-local stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established.

@@

Solving this equation yields the familiar Gaussian function, 
\begin{equation}
  s(t) = A e^{ - \beta t^2 / 2},
\end{equation}
where $\beta$ can be interpreted as the inverse variance or dispersion of the growth phenomena and where $A=1$ when $s(0)=1$. Thus, not surprisingly Gaussian growth emerges for mutually independent events that have a linear time dependence. The same result was originally formulated by Gauss with space as the independent variable instead of time and with error as the dependent variable instead of abundance, all in the context of the prediction of the heavenly object called Ceres \cite{gauss1857theory} (see \cite{stahl2006evolution} for details). In the context of measurement error, one would equate the above equation to the statement that the relative measurement error decreases linearly with size.

@@

For many growth processes within biological organisms, it has been well established that the mediator is the electromagnetic field \cite{becker1984electromagnetic}, which suggests that the accompanying growth pattern in these processes follows the Gompertz model as the result of a ubiquitous growth-inducing field. For growth in larger ecological or societal systems other catalysts can be air pollution, oxygen deprivation, or even instant psychologically induced catalysts.

@@

% \section*{Author Contributions}
% M. A. Haugen: Conceptualization, Methodology, Software, Writing. D. Gilbert: Conceptualization, Reviewing and Editing.

\section{Appendix A}
\label{appendix:a}
Here we arrive at a closed-form expression for a finite number of interaction terms in this modified logistic model, which represents the infection term in the SIR model. Working with a single unitless growth variable and omitting the multivariate network without loss of generality, we start with the logistic growth equation augmented by $K$ terms and using the nomenclature defined in Section \ref{sec:micro}, viz.

@@

>>>>

It is worth noting that the notion that Gompertz growth is maximally coherent and non-local has also been derived from quantum mechanical principles \cite{molski2003coherent}. Molski and Konarski (2003) showed that the Gompertz equation is the solution of a form of the time-independent Schrodinger equation. From this result, they derive the quantum coherent states which are non-local in space while moving along a classical time trajectory. This adds to our exhibits that the Gompertz model can be interpreted as non-local in origin. Coherent quantum states was originally formulated in the context of electromagnetic fields by \citet{glauber1963coherent} (for which he received the Nobel Prize). Some biological macroscopic growth phenomena are also due to such fields. For example, the role of internal electrical fields in regenerative growth is well established and is based on negative electrical potentials that act on sensitive cells \cite{becker1984electromagnetic}.

@@

On the topic of disease modelling, the Gompertz model has been shown by Wang et al. \cite{wang2012richards} to be excluded from a subset of models in the communicable disease paradigm. They used the common Susceptible-Infected-Recovered (SIR) model by \citet{kermack1927contribution} augmented by the Richards parameter and showed that the allowed parameter range with the communicable disease assumption does not cover the parameter value taken in the Gompertz scenario. If one incorporates a social proximity network on which the disease spreads, \citet{zonta2022virus} observed through computer simulations that a near-Gompertz growth can occur but only for scale-free networks, a type of network that has been shown to be rare in human social networks \cite{clauset2009power,broido2019scale,holme2019rare}. 
We will show that one can obtain a true Gompertz curve from a ubiquitous stressor scenario based on microscopic interactions.
% Given these observations, any mortality pattern that shows a Gompertz characteristic seems not be a result of a communicable disease, but rather from a ubiquitous stressor. 
While many studies show the SIR model (and its extensions) as the fitted model in epidemics \cite{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}, some recent studies have successfully fitted a Gompertz model \cite{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}.

\subsection{Main Results and Roadmap}
In this paper, our primary aim is to investigate the source of growth in self-organized processes. Second, we will elaborate on the idea that the logistic model results from local interactions while the Gompertz model results from a field or equivalently from an infinite set of higher order interactions between the microscopic entities involved in the growth. We will devise a slightly different parametrization than the Richards model, the generalized Gompertz-Logistic equation which is commonly used to compare the two models (e.g. by \citet{petroni2020logistic}, \citet{tjorve2017use}, and \citet{wang2012richards}). In particular, instead of considering a decimal number as the unifying exponent we will appeal to integer exponents with a number of terms that tend towards infinity as we approach the Gompertz model. As we shall see, this is rooted in the microscopic interpretation of the growth system, where each subgroup of $N$ elementary entities interacts with all other equally sized subgroups each composed of mutually exclusive elementary entities. If these subgroups are present at all orders of cardinality, the Gompertz model emerges, while only pairwise interactions yield the logistic model. As seen in the aforementioned references, this is in line both with experimental and theoretical results. In presenting this argument, we present a novel bridge between the logistic and the Gompertz model which offers both a mathematical and biological foundation to the source of growth. Lastly, we will present a unifying framework in the growth context for the common functions Gaussian, Exponential, Logistic, Gompertz and Poisson.

Our paper is structured as follows: we will first present a new macroscopic foundation of the Gompertz-Logistic spectrum which offers an alternative to the traditional Richards model. Then, we present a microscopic foundation which acts as the building blocks of the macroscopic phenomena. This is where we show that the Gompertz model results from an infinite set of higher order interactions between the elementary entities involved in the growth, which could alternatively be seen as a ubiquitous field or in the biological context: an environmental stressor. We proceed with a discussion that summarizes our findings in the context of the other well-known growth modalities involving the Gaussian and the Laplacian model and then conclude.

@@

Consider $N$ entities represented by a set of random abundance variables, $x_i(t), \quad i=1, ..., N$, which are unitless and take values between 0 and 1, where 0 represents no realized growth while 1 represents the maximum potential growth having been realized. In other words, these variables represent the \textit{realized potential}, either in the form of individual volume, mass, likelihood of infection or mortality by being preyed on, etc. 
Let these variables have a shared expected value as a function of time, $\mu(t)$, and corresponding higher moments, i.e. each variable follows the same statistical distribution.
With these variables it will also be convenient to also define an \textit{abundance potential} variable, $s_i = 1 - x_i$, where we omit the temporal variable for ease of exposition. An abundance potential could represent the future capacity to grow, e.g. encoded in an organism's DNA combined with its environmental factors. Our goal now is to investigate under which conditions we can achieve logistic or Gompertz growth.

Preliminarily, consider a network between these $N$ entities along a graph with a set of edges and nodes which delineates the interaction of the growth variables, where we define interaction in this context as the transfer of one entity's potential growth to another entity which converts potential to realized growth. Examples of such networks could be cells that communicate either through electron or protein transfer, predator-prey systems or even human interactions through mutual trade, spreading of knowledge, or disease.

\subsection{Microscopic Logistic Growth}

Now consider the growth equation with only pairwise interactions, or \textit{first order} interactions, governed by an adjacency matrix ${a_{ij}}$ as a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise, viz.
\begin{equation}
\label{eq:modSIR}
\frac{d x_i}{dt} = x_i\sum_{j \in \mathcal{N}(i)}{a_{ij}}s_{j} \quad \forall i,
\end{equation}
or
\begin{equation}
\label{eq:modSIR2}
\frac{d}{dt}\ln x_i = \sum_{j \in \mathcal{N}(i)}{a_{ij}}s_{j} \quad \forall i,
\end{equation}
where $\mathcal{N}(i)$ is the neighborhood of nodes connected to node $i$. For example, if the adjacency matrix has only $1$s and no $0$s, this means that every node is connected to every other node. The diagonal of the adjacency matrix is zero as a node is not connected to itself by definition.

Let us look at the simplest case of two connected entities only, viz.
\begin{eqnarray}
\frac{d x_1}{dt} = x_1 s_2, \\
\frac{d x_2}{dt} = x_2 s_1.
\end{eqnarray}
Rearranging and summing the two equations, we can express the variables in terms of their averages,
\begin{equation}
\frac{1}{2} \left\{\frac{1}{x_1}\frac{d x_1}{dt} + \frac{1}{x_2}\frac{d x_2}{dt}\right\} = \frac{1}{2}(s_1 + s_2) = 1 - \frac{1}{2}(x_1 + x_2),
\end{equation}
where the left hand side represents the average of the relative growth, while the right hand side represents the average abundance potential. Thus, even for a simple two-variable system we see that the system can be described in terms of summary statistics, which we shall see that we will be able to generalize.

We choose to work in a unitless domain without parameters for the interaction strength and carrying capacity for simplicity of exposition. These can be added without changing the conclusions of these arguments.

Now as we let N grow, we impose a simple rule: \textit{Each entity can only receive growth potential from a single entity and create growth potential in a single possibly different entity}. This condition translates into an adjacency matrix with each column containing a single non-zero entry. 
Using this rule and rearranging and summing across all entities we obtain the following relationship,
\begin{equation}
\label{eq:modSIR2}
\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ] = \sum_j^N s_{j}.
\end{equation}

Dividing by the total number of entities, we obtain the simplest form of logistic growth of the geometric mean of the ensemble driven by the arithmetic mean of the inverse abundance, or the \textit{potential abundance}, 
\begin{equation}
\label{eq:modSIR3}
\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} = \frac{1}{N}\sum_j^N s_j.
\end{equation}

Now note that the time derivative of the geometric mean translates into the \textit{average relative growth rate}, where if we let $v_i=\frac{1}{x_i}\frac{d}{dt}x_i$, we see that

\begin{equation}
\label{eq:modSIR4}
\frac{1}{N}\sum_j^N v_i = \frac{1}{N}\sum_j^N s_j.
\end{equation}

At this point we have shown how macroscopic variables can be obtained from their microscopic counterparts. We have not needed to evoke the randomness of our variables, but have obtained a macroscopic understanding only using sample quantities starting from pairwise microscopic principles.

However, if we assume our set of random variables, $\{x_i\}$, converge to a distributional mean as $N$ grows, both of the relative growth rate average and the abundance average will converge to their distributional counterparts. Specifically and with a slight abuse of notation in the arguments, letting $\bar{x}(N)$ represent the sample abundance mean for N nodes which converges to its distributional counterpart $\mu(t)$, it follows from the continuous mapping theorem that $\bar{v}(N)$ will converge to $\frac{1}{\mu(t)}\frac{1}{dt}\mu(t)$ so long as the time derivative is smooth. We thus have
\begin{eqnarray*}
  \lim_{N\rightarrow \infty} \frac{1}{N}\sum_j^N v_i &=& \lim_{N\rightarrow \infty} \frac{1}{N}\sum_i^N s_i \\\
  \lim_{N\rightarrow \infty} \bar{v}(N) &=& \lim_{N\rightarrow \infty}(1 - \bar{x}(N)) \\
   \frac{1}{\mu(t)}\frac{d}{dt}\mu(t) &=&1 - \mu(t),
\end{eqnarray*}
where we make the time dependency explicit for clarity of exposition. Now we see that we have obtained the logistic model in terms of a macroscopic ensemble average.

We make special note that logistic growth emerges from pairwise interactions only and without any independence assumption across the variables. This does not mean that the logistic growth cannot emerge from other microscopic settings but it nevertheless provides one possible scenario under which the logistic curve can emerge. Now, as we shall allow higher order interactions we will see the Gompertz curve emerge.

@@

Thus, let $S_k$ be the set of all mutually exclusive $k$-sized groups where $k=1,..., N/2$, where the highest order are capped at $N/2$ since that would be the largest number of entities in a single group that could interact with an identically sized group. Thus, the aggregated growth is represented by the sum of each interaction order,
\begin{equation}
\label{eq:MicroGomp0}
\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ] = \sum_{j\in S_1}^N s_{j} + \sum_{j_1, j_2 \in S_2} s_{j_1}s_{j_2} + ... + \sum_{j_1, j_2, ..., j_{N/2} \in S_{N/2}} s_{j_1}s_{j_2}\dots s_{j_{N/2}},
\end{equation}
where $S_k$ is the set of all mutually exclusive $k-$sized groups. In each of the sums, the total number of terms is equal to the total number of groups which at each order is $N/k$ where $k$ is the interaction order.

Now, we can use the identical distribution assumption and divide the right hand side by $N$ to identify convergence properties as both $k$ and $N$ grow:
\begin{enumerate}
\item As the number of entities grow each term will converge in expectation to higher powers of the mean of $s_i$, multiplied by the prefactor $1$, $1/2$, $1/3$, etc., 
\begin{equation}
\label{eq:MicroGomp11}
\lim_{N\rightarrow \infty} \frac{1}{N}\sum_{j_1, j_2, ..., j_k \in S_k} s_{j_1}s_{j_2}\dots s_{j_k} = \frac{1}{N}\frac{N}{k} \mu^{k} = \frac{1}{k} \mu^{k}.
\end{equation}
\item As the higher order interaction terms grow
\begin{equation}
\label{eq:MicroGomp12}
\lim_{k\rightarrow \infty} \frac{1}{N}\sum_{j_1, j_2, ..., j_k \in S_k} s_{j_1}s_{j_2}\dots s_{j_k} = 0
\end{equation}
\end{enumerate}
Thus, as each higher order term is composed of a decreasing number of summations and an increasing number of multiplications, the product of $s_i$ goes to zero since each random variable is bounded between 0 and 1. 
Now, using the Taylor series identity $\mu+\mu^2/2+... = -\ln(1-\mu)$, we obtain Gompertz growth in terms of the arithmetic means of the microscopic entities
\begin{eqnarray}
\label{eq:MicroGomp13}
\lim_{N\rightarrow \infty}\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ]^{1/N} &=&  \lim_{N\rightarrow \infty}\left[\sum_j^N s_{j} + \sum_{j_1, j_2 \in S_2} s_{j_1}s_{j_2} + ...\right]\\
\lim_{N\rightarrow \infty} \bar{v}(N)  &=&  \mu + \frac{1}{2} \mu^2 + ...\\
\frac{1}{\mu(t)}\frac{d}{dt}\mu(t) &=& - \ln (1-\mu). 
\end{eqnarray}

Note that for the right hand side to exhibit asymptotically infinite growth rate at the onset of growth, it is not only sufficient but also necessary to have higher order terms. This is seen by virtue of the infinite polynomial representation of the logarithm and implies that \textit{Gompertz growth cannot result from pairwise interactions under the assumptions presented in this model}, but rather requires all higher order interactions to be present.

Since all higher order interaction would necessarily imply non-local interaction, the natural interpretation would be that the growth is activated by a ubiquitous field accessible by all entities simultaneously. 
In this way one could think of the activating field as an \textit{effectively} external field, even though it is possible that the field emerges as a collective emergent phenomenon purely from within the system or organism. Examples of this field could be sound waves or the electromagnetic field.

Another way to see the source of growth as effectively external is to recognize that the source of growth in the higher order terms is coming from the abundance potential and not the realized potential, while in the logistic growth model this is not necessarily the case. 
To see this, first swap the role of $x_i$ and $s_i$ on the right hand side of Equation \ref{eq:modSIR} to see that the same logistic dependency can be obtained under the exclusivity condition. This could be interpreted as a causality reversal of the growth process.
Then note that when augmenting the growth equation with higher order terms on the right-hand side of the same equation, the same swapping will not lead to the Gompertz model (see Appendices \ref{appendix:swappinglogistic} and \ref{appendix:swappinggompertz}).

Causality reversal is indeed what is done when modelling communicable diseases in a microscopic network context, where $x_i$ represents the infected nodes and $s_i$ represents the susceptible nodes (see e.g. \citet{estrada2022networked}). But because higher order growth does not allow this causality reversal, where the infected nodes are the source of spread, the Gompertz model cannot be used since it would imply that the susceptible nodes are the source of the disease spreading, which would turn communicable disease theory on its head. In this context, the Gompertz model would rather suggest a system disturbed by a ubiquitous, simultaneous, and non-local stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established. Thus, with the necessary requirement of higher order interactions between nodes, \textit{the Gompertz model cannot be used as a communicable disease model}.
% While some applied work show how scale-free pairwise interactions approximate Gompertz, theoretical parity is not shown (e.g. \citet{zonta2022virus}).

@@

Solving this equation yields the familiar Gaussian function, 
\begin{equation}
  s(t) = A e^{ - \beta t^2 / 2},
\end{equation}
where $\beta$ can be interpreted as the inverse variance or dispersion of the growth phenomena and where $A=1$ when $s(0)=1$. Thus, not surprisingly Gaussian growth emerges for mutually independent events that have a linear time dependence. The same result was originally formulated by Gauss with space as the independent variable instead of time and with spatial estimation error as the dependent variable instead of abundance, all in the context of the prediction of the heavenly object called Ceres \cite{gauss1857theory} (see \cite{stahl2006evolution} for details). In the context of measurement error, one would equate the above equation to the statement that the relative measurement error decreases linearly with size.

@@

For many growth processes within biological organisms, it has been well established that the mediator is the electromagnetic field \cite{becker1984electromagnetic}, which suggests that the accompanying growth pattern in these processes follows the Gompertz model as the result of a ubiquitous growth-inducing field. For growth in larger ecological or societal systems other catalysts can be air pollution, oxygen deprivation, or even instant psychologically induced catalysts like collective fear.

@@

% \section*{Author Contributions}
% M. A. Haugen: Conceptualization, Methodology, Software, Writing. D. Gilbert: Conceptualization, Reviewing and Editing.
\appendix
\section*{Appendices}
\addcontentsline{toc}{section}{Appendices}
\renewcommand{\thesubsection}{\Alph{subsection}}
\subsection{A closed-form finite-interaction model}
\label{appendix:a}
Here we arrive at a closed-form expression for a finite number of interaction terms in this modified logistic model, which represents the infection term in the SIR model. Working with a single unitless growth variable and omitting the multivariate network without loss of generality, we start with the logistic growth equation augmented by $K$ terms and using the nomenclature defined in Section \ref{sec:micro}, viz.

@@

\subsection{The invariance of logistic growth under causality reversal}
\label{appendix:swappinglogistic}
Consider a causality reversal of logistic growth such that the realized growth drives potential growth. Then Equation \ref{eq:modSIR} must be modified by swapping $x$ and $s$ on the RHS as follows,
\begin{equation}
\label{eq:revSIR}
\frac{d x_i}{dt} = s_i\sum_{j \in \mathcal{N}(i)}{a_{ij}}x_{j} \quad \forall i.
\end{equation}
Using the identity,
\begin{equation}
\label{eq:modSIR2}
\sum_i\frac{1}{s_i}\frac{dx_i}{dt} = -\sum_i\frac{1}{s_i}\frac{ds_i}{dt} = -\frac{d}{dt} \ln \left[ \prod_i^N s_i \right ],
\end{equation}
and using the equivalent connectivity rule as outlined in Section \ref{sec:micro}, we have,
\begin{equation}
\label{eq:rev_id}
-\frac{d}{dt} \ln \left[ \prod_i^N s_i \right ] = \sum_j^N x_j.
\end{equation}

And using the same convergence argument as before we get $\bar{s}(N) \rightarrow \gamma(t) = 1 - \mu(t)$
\begin{equation}
  -\frac{1}{\gamma(t)}\frac{d}{dt}\gamma(t) = (1-\gamma(t)),
\end{equation}
which solves to 
\begin{equation}
  \gamma(t) = \frac{1}{1 + A \exp(t)},
\end{equation}
which implies that
\begin{equation}
  \mu(t) = \frac{A \exp(t)}{1 + A \exp(t)}.
\end{equation}
So even when realized growth is responsible for growth thus reversing causality in the pairwise growth setting, the logistic emerges. Thus, we can say that the \textit{microscopic logistic model is invariant under causality reversal}.

\subsection{The lack of invariance of Gompertz growth under causality reversal}
\label{appendix:swappinggompertz}
Similar to our procedure with the logistic model we now swap potential growth with actual growth on the RHS of the microscopic Gompertz model to simulate a causality reversal,
\begin{equation}
\label{eq:MicroGompRev}
-\frac{d}{dt} \ln \left[ \prod_i^N s_i \right ] = \sum_{j\in S_1}^N x_{j} + \sum_{j_1, j_2 \in S_2} x_{j_1}x_{j_2} + ... + \sum_{j_1, j_2, ..., j_{N/2} \in S_{N/2}} x_{j_1}x_{j_2}\dots x_{j_{N/2}}.
\end{equation}
Make a special note of the new minus sign on the LHS due to a change of variable as done with the logistic treatment (Eq. \ref{eq:rev_id}).

Now, with our continuity argument as outlined in the main text we obtain the macroscopic version of this equation as
\begin{equation}
  \frac{d}{dt}\ln \gamma(t) = \ln \gamma(t),
\end{equation}
with $\gamma(t) = 1-\mu(t)$ and where we are now rid of the minus sign on the RHS of the regular Gompertz model (Eq. \ref{eq:Gomp} due to causality reversal.

This equation solves to
\begin{equation}
  \mu(t) = 1 - \exp\left[ A\exp(t)\right],
\end{equation}
which is not of the Gompertz form and furthermore does not exhibit the initially fast growth which is characteristic of the Gompertz model (see Figure \ref{fig:causality_reversal}). Thus we have shown that the \textit{microscopic Gompertz model is not invariant under causality reversal} and is thus not suitable for a communicable disease setting where infected nodes infect uninfected nodes.

\insertSmallPdfFig{causality_reversal}{A comparison of the growth profiles of the Gompertz and the causality-reversed Gompertz models. Their relative growth rates defined as the time derivative of the logarithm of the abundance ($\mu(t)$) are also shown. To solve for the constants involved we apply the initial condition $\mu(0) = 0.001$ (a very small positive number).}
